---
title: "RMF Analysis"
author: "Group 2"
output: html_notebook
---

```{r}
load(url("http://statmath.wu.ac.at/~vana/datasets/RMFAnalysisClean.rda"))
```

```{r}
if (!require("scatterplot3d")) install.packages("scatterplot3d"); library(scatterplot3d)
if (!require("cluster")) install.packages("cluster"); library("cluster")
if (!require("dbscan")) install.packages("dbscan"); library("dbscan");
```

```{r}
head(RMFAnalysisClean)
```


```{r}
RMFAnalysisClean <- na.omit(RMFAnalysisClean)
```

```{r}
plot(RMFAnalysisClean)
```

```{r}
scatterplot3d(RMFAnalysisClean[,1:3])
```

```{r}
model_scaled <- scale(RMFAnalysisClean)
```

```{r}
plot(model_scaled)
#After scaling we see that the first two principal components are "Monetary.mean" and "Frequency"
```


## K-means
```{r}
# Manually select number of initial values 
nstart <- 5
# Manually select optimal number of clusters 
clusters <- 4
# k-means Clustering
model.pc <- kmeans(model_scaled, 
                   centers = clusters, 
                   nstart = nstart)
```


```{r}
model.pc$centers
```

```{r}
#Here we can see the index of the cluster assigned by `kmeans()` to each observation
model.pc$cluster
```

```{r}
#Here we use the clustering to colour code the points in the scatterplot 
#(cbbPalette is where we put the colours we want to use)
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
plot(model_scaled, col = cbbPalette[model.pc$cluster])
points(model.pc$centers, pch = 3, cex = 2)
text(model.pc$centers, labels = 1:4, pos = 2)
```

## Hierarchical clustering
```{r}
model_dist <- dist(model_scaled) # Euclidean distance matrix
model_hc <- hclust(model_dist)
```

```{r}
#The dendrogram can be visualized by:
plot(model_hc)
```

```{r}
groups <- cutree(model_hc, h = 2)
groups
```

```{r}
groups <- cutree(model_hc, k=clusters)
groups
```

```{r}
plot(model_hc)
rect.hclust(model_hc, k=clusters)
```

```{r}
dif <- par(mfrow = c(1,2))
plot(model_scaled, col=cbbPalette[model.pc$cluster], 
     main = "K-means")
plot(model_scaled, col=cbbPalette[groups], 
     main = "hclust")
par(dif)
```

## DBSCAN
```{r fig.height=6, fig.width=6}
### Partitional Clustering (k-Means)
clusters <- 3
model_km <- kmeans(scale(RMFAnalysisClean), 
                   clusters)
plot(RMFAnalysisClean, col = cbbPalette[model_km$cluster])
```

```{r fig.height=8, fig.width=10}
### DBSCAN - minPts = roughly dimension+1
pal <- c("#000000","#004949","#009292","#ff6db6","#ffb6db",
 "#490092","#006ddb","#b66dff","#6db6ff","#b6dbff",
 "#920000","#924900","#db6d00","#24ff24","#ffff6d")
model_dbs <- dbscan::dbscan(RMFAnalysisClean, 
                            eps = 7, 
                            minPts = 3)
plot(RMFAnalysisClean, col = pal[model_dbs$cluster + 1])
```

```{r}
model_dbs
```

```{r}
kNNdistplot(RMFAnalysisClean, k = 3) 
abline(h = 6, lty = 2)
```

```{r fig.height=8, fig.width=10}
### DBSCAN - minPts = roughly dimension+1
model_dbs2 <- dbscan::dbscan(RMFAnalysisClean, 
                            eps = 5.5, 
                            minPts = 3)
plot(RMFAnalysisClean, col = pal[model_dbs2$cluster + 1])
```

```{r fig.height=8, fig.width=10}
### Hierarchical Clustering
md <- dist(scale(RMFAnalysisClean))
model_hc <- hclust(md)
plot(RMFAnalysisClean, col=cbbPalette[cutree(model_hc, k=clusters)])
```

```{r}
kNNdistplot(RMFAnalysisClean, k = 3) 
abline(h = 11, lty=2)
```

### Task 3: final suggestion

The DBSCAN algorithm does not work as the eps that is found using the elbow method is too big to find meaningful clusters within the data. This is likely because the data has clusters with varying densities and is higher dimensional.
This leaves us with the k-means and hierarchical clustering. The k-means algorithm partitions the data into roughly equal size clusters. The hierarchical clustering on the other hand clusters the data into a big cluster near the origin and then a cluster of outliers along each axis.
As the dataset seems to be rather homogeneous, we believe that the hierarchical clustering makes the most sense for an online retailer as this means that frequent shoppers who do not spend much on average per buy are clustered as are non-frequent shoppers who tend to spend a lot per buy for the outlier clusters. This would enable the online store to target these segments with specific ads and the like.
